%!TEX root = ../main.tex
\section{Proposed Method}
\label{sec:method}

\subsection{Overview}

We propose a language-conditioned trajectory diffusion model for whole-body humanoid control. 
The pipeline consists of three stages:

\begin{enumerate}
    \item \textbf{Pretraining:} Train an unconditional diffusion model $p_\theta(\tau)$ over humanoid motion trajectories.
    \item \textbf{Conditioning:} Introduce text conditioning $p_\theta(\tau \mid z_{\text{text}})$ using cross-attention or FiLM layers.
    \item \textbf{Control Integration:} Deploy generated trajectories into a PD controller for stable execution
\end{enumerate}

This hybrid approach combines imitation learning with policy optimization for robust, semantically aligned control.

\subsection{System Architecture}

\begin{itemize}
    \item \textbf{Inputs:} text prompt and proprioceptive history.
    \item \textbf{Outputs:} time-parameterized per-joint control trajectories.
    \item \textbf{Backbone:} shared temporal encoder with decoders for motion and language.
    \item \textbf{Losses:} diffusion loss, imitation loss, and alignment loss between text and motion.
\end{itemize}

\subsection{Data and Tokenization}

We utilize BeyondMimic \cite{liao2025beyondmimicmotiontrackingversatile} and LAFAN1 \cite{harvey2020robust} datasets for pretraining motion representations. 
A spatiotemporal tokenizer aligns motion frames with textual segments, and time normalization is applied to enhance temporal generalization.

\subsection{Evaluation Plan}

\textbf{Baselines:}
\begin{itemize}
    \item BeyondMimic RL (task-specific policy)
    \item ExBody2 imitation controller
    \item Diffusion model without language conditioning
\end{itemize}

\textbf{Metrics:}
\begin{itemize}
    \item Task success rate (prompt compliance)
    \item Motion realism (FID in latent space)
    \item Trajectory smoothness and stability
    \item Language–motion alignment (CLIP similarity or human evaluation)
\end{itemize}

\textbf{Expected Outcomes:}
\begin{itemize}
    \item Demonstrate feasibility of language-conditioned humanoid control.
    \item Release a dataset of captioned humanoid trajectories.
    \item Provide an open-source simulation and training pipeline.
\end{itemize}